---
title: "Antarctic Marine Species Traits"
subtitle: "A hands-on tour of the `worrms` R package"
author: "Pablo Deschepper | Institute for Natural Sciences, Brussels"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, metropolis, metropolis-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: "%current% / %total%"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.retina = 2,
  fig.align = "center",
  out.width = "80%"
)
```

```{r packages, include=FALSE}
library(worrms)
library(tidyverse)
library(rgbif)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(knitr)
library(kableExtra)
library(qrcode)
```

---
class: inverse, center, middle

# What is WoRMS?

---

# World Register of Marine Species

.pull-left[
The **authoritative global checklist** of all known marine species.  
Maintained by taxonomic experts, hosted by **VLIZ**.

**WoRMS provides:**
- Validated accepted names + full synonymy
- Taxonomic classification (kingdom → species)
- Trait / attribute data (functional groups, body size, IUCN status)
- Geographic distribution
- Cross-links to GBIF, BOLD, FishBase, IUCN…

🌐 [www.marinespecies.org](https://www.marinespecies.org)
]

.pull-right[
<br><br>
**The `worrms` R package**

- Official rOpenSci R client for the WoRMS REST API
- No API key required
- Every function starts with `wm_`
- Functions ending in `_` are **vectorised**
- All results returned as **tibbles**

```r
install.packages("worrms")
library(worrms)
```
]


---

# Every taxon has a unique AphiaID

The **AphiaID** is a stable integer identifier — the primary key across all `worrms` functions.

.pull-left[
```{r aphia1, eval=FALSE}
# Name → AphiaID
krill_id <- wm_name2id("Euphausia superba")
krill_id
```
```{r aphia1-out, echo=FALSE}
236217
```

```{r aphia2, eval=FALSE}
# AphiaID → name
wm_id2name(236217)
```
```{r aphia2-out, echo=FALSE}
"Euphausia superba"
```
]

.pull-right[
```{r aphia3, eval=TRUE}
# Vectorised batch lookup (one API call)
spp <- c("Euphausia superba",
         "Dissostichus eleginoides",
         "Notothenia coriiceps")

wm_name2id_(spp)
```
]

---
class: inverse, center, middle

# Function families

---

# Family 1 — Full taxon records

`wm_record()` is the workhorse: ~25 fields per taxon.

.pull-left[
```{r record, eval=TRUE}
krill_record <- wm_record(236217)[1:15] |>
glimpse() 
```
]

.pull-right[
| Field | Meaning |
|---|---|
| `AphiaID` | Unique WoRMS identifier |
| `scientificname` | Name queried |
| `status` | `"accepted"` or `"unaccepted"` |
| `valid_AphiaID` | Points to accepted record |
| `rank` | Species, Genus, … |
| `isMarine` | Habitat flag |
| `isBrackish` | Habitat flag |
| `isFreshwater` | Habitat flag |
| `isExtinct` | Extinction status |
| `lsid` | Globally unique URI |
]

---

# Family 1 — Searching & fuzzy matching

.pull-left[
**Exact & fuzzy search**
```{r records-name, eval=TRUE}
# Exact match
wm_records_name("Euphausia superba") |>
  select(AphiaID, scientificname, status)

# Fuzzy — great for typos
wm_records_name("Euphausia su",
                fuzzy = TRUE) |>
  select(scientificname, match_type)
```
]

.pull-right[
**TAXAMATCH** — WoRMS' own algorithm, ideal for cleaning field species lists

```{r taxamatch, eval=TRUE}
messy <- c("Euphausia superba",  # correct
           "Euphausia supurba",  # typo
           "Notothenia coriceps")# missing letter

res <- wm_records_taxamatch(messy)

bind_rows(res, .id = "query") |>
  select(query, scientificname,
         status, match_type)
```
]

---

# Family 2 — Taxonomy tree navigation

.pull-left[
**Full classification path**
```{r class, eval=TRUE}
wm_classification(236217) |>
  select(rank, scientificname) |> head()
```
]

.pull.right[
**Direct children of a taxon**
```{r children, eval=TRUE}
euphausiacea_id <- wm_name2id(
  "Euphausiacea")

wm_children(euphausiacea_id) |> select(AphiaID, scientificname, status )
```
]


---

# Family 2 — Taxonomy tree navigation

.pull-left[
**Rank lookups**
```{r ranks, eval=FALSE}
wm_ranks_id(220)         # → "Species"
wm_ranks_name("Species") # → 220
```

> ⚠️ The same rank (e.g. *Species*) exists  
> independently in multiple kingdoms —  
> rank IDs are kingdom-specific.
]

.pull.right[
**All species under a parent taxon**
```{r rank-search, eval=TRUE}
euphausiidae_id <- wm_name2id(
  "Euphausiidae")

wm_records_rank(220,
  id = euphausiidae_id) |> select(AphiaID, scientificname, status) |> head()
```
]

---

# Family 3 — Synonyms & vernacular names

.pull-left[
**Synonyms**
```{r syn, eval=TRUE}
# Euphausia superba
syn <- wm_synonyms(236217)
syn |> select(scientificname,
              status)
```
]

.pull-right[
**Common / vernacular names**
```{r common, eval=TRUE}
# All common names for krill
wm_common_id(236217)
```
]

---

# Family 4 — Distribution & external IDs

.pull-left[
**Geographic distribution**
```{r dist, eval=TRUE}
krill_dist <- wm_distribution(236217)

# Records per locality
krill_dist |>
  count(locality, sort = TRUE) |>
  head(10)
```

**Literature sources**
```{r sources, eval=FALSE}
wm_sources(236217) |>
  select(source_id, reference, page)
```
]

.pull-right[
**Cross-links to external databases**

| Code | Database |
|---|---|
| `tsn` | ITIS |
| `bold` | Barcode of Life |
| `eol` | Encyclopedia of Life |
| `col` | Catalogue of Life |
| `iucn` | IUCN Red List |
| `fishbase` | FishBase |

```{r external, eval=TRUE}
wm_external(234700, type = "fishbase")
```
]

---
class: inverse, center, middle

# Family 5 — Trait attributes

---

# Trait attributes — overview

WoRMS stores structured **functional trait data** attached to taxa.

.pull-left[
**Browse attribute type definitions**
```{r attr-def, eval=FALSE}
wm_attr_def(id = 1)   # IUCN Red List
wm_attr_def(id = 7)   # Functional group
wm_attr_def(id = 15)  # Body size
```
]

.pull-right[
**Browse values within a category**
```{r attr-cat, eval=TRUE}
wm_attr_category(id = 7) |>
  select(measurementValueID,
         measurementValue)
```
]

---

# Trait attributes — overview

**Fetch traits for a taxon**
```{r attr-data, eval=TRUE}
krill_attrs <- wm_attr_data(236217)
krill_attrs |> glimpse()
```

---

# Trait attributes — overview

**Vectorised batch retrieval**
```{r attr-batch, eval=TRUE}
ids <- unlist(wm_name2id_(c(
  "Euphausia superba",
  "Dissostichus eleginoides",
  "Notothenia coriiceps")))

bind_rows(wm_attr_data_(ids)) |>
  select(AphiaID, measurementType,
         measurementValue)
```

---

# Trait attributes — querying by attribute

`wm_attr_aphia()` flips the question: *which taxa have a given attribute?*

```{r attr-aphia, eval=TRUE}
# All taxa with an IUCN Red List category (attribute type 1)
# Results are paginated — up to 50 per page
taxa_with_iucn <- wm_attr_aphia(id = 1)
head(taxa_with_iucn, 10)
```
---
class: inverse, center, middle

# Use-case
### Antarctic *Notothenioidei* + GBIF

---

# The workflow

.center[
```
 WoRMS (worrms)                      GBIF (rgbif)
 ─────────────────────────           ──────────────────────────
 wm_name2id("Notothenioidei")
 wm_children()                ──►    occ_search() south of 60°S
 wm_records_rank()            ──►    bind_rows() + filter coords
                                     ↓
                                  visualise (ggplot2 + sf, EPSG:3031)
```
]

**Goal:** map *Notothenioidei* (dominant Antarctic fish suborder) occurrences  
with **colour = family** and explore **species richness** and **sampling effort** through time.

---

# Step 1 — Build the species list

.pull-left[
```{r usecase-subtree, eval=TRUE}
notothenioidei_id <- wm_name2id("Notothenioidei")

# Confirm taxonomic position
wm_classification(notothenioidei_id) |> 
  select(rank, scientificname)
```
]

.pull-right[
```{r usecase-families, eval=TRUE}
# Direct children (families)
notothenioidei_families <- tryCatch(
  wm_children(notothenioidei_id),
  error = function(e) { message(conditionMessage(e)); tibble() }
)

# All accepted species per family (rank 220 = Species)
family_ids <- filter(notothenioidei_families, rank == "Family")$AphiaID
family_names <- filter(notothenioidei_families, rank == "Family")$scientificname

species_list <- map(family_ids, \(fid)
  tryCatch(wm_records_rank(220, id = fid), error = \(e) NULL))

names(species_list) <- family_names

notothenioidei_spp <- bind_rows(species_list, .id = "family") |>
  filter(status == "accepted")

cat("Total accepted species:", nrow(notothenioidei_spp))
```
]
---

# Step 2 — GBIF occurrences south of 60°S

```{r usecase-gbif, eval=TRUE}
target_species <- notothenioidei_spp$scientificname

gbif_results <- map(target_species, \(sp)
  tryCatch(
    occ_search(
      scientificName   = sp,
      decimalLatitude  = "-90,-60",
      hasCoordinate    = TRUE,
      limit            = 200,
      fields           = c("name", "decimalLatitude", "decimalLongitude",
                           "year", "datasetName", "species")
    )$data,
    error = \(e) {
      message("Error for ", sp, ": ", conditionMessage(e))
      NULL
    }
  ))

gbif_df <- bind_rows(gbif_results) |>
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude))

cat("Total GBIF records:", nrow(gbif_df))
```

---

# Step 3 — Join gbif data with taxonomy

```{r usecase-join, eval=TRUE}
# WoRMS lookup table
worms_lookup <- notothenioidei_spp |>
  filter(scientificname %in% target_species) |>
  select(scientificname, AphiaID, family, isMarine, isExtinct)

# Join onto GBIF records
gbif_enriched <- gbif_df |>
  rename(scientificname = species) |>
  left_join(worms_lookup, by = "scientificname")
```

---

# Step 4 — Occurrence map (EPSG:3031)

```{r usecase-map, eval=TRUE, fig.width=8, fig.height=4.5}
world   <- ne_countries(scale = "medium", returnclass = "sf")
gbif_sf <- st_as_sf(gbif_enriched,
                    coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

occplot <- ggplot() +
  geom_sf(data = st_transform(world, 3031),
          fill = "grey80", colour = "white", linewidth = 0.15) +
  geom_sf(data = st_transform(gbif_sf, 3031),
          aes(colour = family), alpha = 0.7, size = 1.8) +
  coord_sf(crs = 3031, xlim = c(-4e6, 4e6), ylim = c(-4e6, 4e6)) +
  scale_colour_brewer(palette = "Set3", name = "Family") +
  labs(title    = "Notothenioidei — GBIF occurrences south of 60°S",
       subtitle = paste0(nrow(gbif_enriched), " records | ",
                         n_distinct(gbif_enriched$scientificname), " species"),
       x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))
```

---

# Step 4 — Occurrence map (EPSG:3031)

```{r occurrence map, eval=TRUE, fig.width=8, fig.height=4.5}
occplot

```

---

# Step 5 — Species richness & sampling effort

.pull-left[
```{r richness-plot, eval=TRUE}
richness_plot <- notothenioidei_spp |>
  count(family, name = "n_species") |>
  mutate(family = fct_reorder(
    family, n_species)) |>
  ggplot(aes(n_species, family,
             fill = n_species)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_c(
    option = "mako", direction = -1) +
  labs(title = "Species richness",
       x = "Accepted species", y = NULL) +
  theme_minimal()
```
]

.pull-right[
```{r richness-plotted, eval=TRUE}
richness_plot
```
]

---

# Step 5 — Species richness & sampling effort

.pull-left[
```{r temporal-plot, eval=TRUE}
temporal_plot <- gbif_enriched |>
  filter(!is.na(year)) |>
  mutate(decade = (year %/% 10) * 10) |>
  count(decade, family) |>
  ggplot(aes(decade, n, fill = family)) +
  geom_col() +
  scale_fill_brewer(
    palette = "Set3", name = "Family") +
  labs(title = "Records per decade",
       x = "Decade",
       y = "N occurrences") +
  theme_minimal() +
  theme(legend.position = "bottom")
```
]

.pull-right[
```{r temporal-plotted, eval=TRUE}
temporal_plot
```
]

---
class: inverse, center, middle

# Quick reference & best practices

---

# All `worrms` functions at a glance

.pull-left[
| Function | Purpose | `_` variant? |
|---|---|:---:|
| `wm_name2id()` | Name → AphiaID | ✓ |
| `wm_id2name()` | AphiaID → name | ✓ |
| `wm_record()` | Full AphiaRecord | ✓ |
| `wm_records_name()` | Search by name | — |
| `wm_records_names()` | Multiple names | — |
| `wm_records_taxamatch()` | TAXAMATCH | — |
| `wm_records_common()` | Search by vernacular | ✓ |
| `wm_records_rank()` | Taxa by rank | — |
| `wm_records_date()` | Modified since date | — |
| `wm_classification()` | Full taxonomy path | ✓ |
| `wm_children()` | Direct children | ✓ |
]

.pull-right[
| Function | Purpose | `_` variant? |
|---|---|:---:|
| `wm_synonyms()` | All synonyms | ✓ |
| `wm_common_id()` | Vernacular names | ✓ |
| `wm_distribution()` | Distribution records | ✓ |
| `wm_external()` | External DB IDs | ✓ |
| `wm_record_by_external()` | External ID → WoRMS | ✓ |
| `wm_sources()` | Literature sources | ✓ |
| `wm_attr_def()` | Attribute definitions | ✓ |
| `wm_attr_category()` | Attribute values | ✓ |
]

---
| Function | Purpose | `_` variant? |
|---|---|:---:|
| `wm_attr_data()` | **Trait data** | ✓ |
| `wm_attr_aphia()` | Taxa by attribute | ✓ |
| `wm_ranks_id/name()` | Rank lookup | — |

---

# Best practices

.pull-left[
**1. Use `_` variants for multiple taxa**
```r
# ❌ N separate API calls
map(my_ids, wm_record)

# ✅ One API call
wm_record_(my_ids)
```

**2. Guard `wm_children()` with `tryCatch()`**
```r
# HTTP 204 = no children → hard error
tryCatch(
  wm_children(id),
  error = function(e) tibble()
)
```
]

.pull-right[
**3. Always check `status`**
```r
records |> filter(status == "accepted")

# Or redirect synonyms:
records |> mutate(
  use_id = if_else(
    status == "accepted",
    AphiaID, valid_AphiaID))
```

**4. Cache during development**
```r
knitr::opts_chunk$set(cache = TRUE)
```

**5. Cite WoRMS**
> WoRMS Editorial Board (2025).  
> World Register of Marine Species.  
> DOI: [10.14284/170](https://doi.org/10.14284/170)
]

---
class: inverse, center, middle

# Thank you

**Full annotated tutorial:**
```{r qr-code, echo=FALSE, fig.width=3, fig.height=3, out.width="25%"}
library(qrcode)
par(mar = c(0, 0, 0, 0))
plot(qr_code("https://thesnakeguy.github.io/worrms_tutorial/worrms_antarctic_tutorial.html"))
```

**Other links:**  
📦 [docs.ropensci.org/worrms](https://docs.ropensci.org/worrms)  
🌐 [www.marinespecies.org](https://www.marinespecies.org)  
🔬 [obis.org](https://obis.org) 
📊 [docs.ropensci.org/rgbif](https://docs.ropensci.org/rgbif)
