---
title: "Exploring Antarctic Marine Species with `worrms`"
subtitle: "World Register of Marine Species R Client"
author: "Pablo Deschepper | Institute for Natural Sciences, Brussels"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  cache     = FALSE,   # set TRUE to cache API calls during development
  fig.width = 9,
  fig.height = 6,
  comment   = "#>"
)
```

---

# Introduction

## What is WoRMS?

The **World Register of Marine Species (WoRMS)** is the authoritative global checklist of all
known marine species. It is maintained by a network of taxonomic experts, hosted by VLIZ and provides:

- Validated, accepted scientific names with full synonymy
- Taxonomic classification (kingdom → species)
- Trait / attribute data (functional groups, IUCN status, habitat)
- Geographic distribution 
- Cross-links to external databases (GBIF, BOLD, …)

**Website:** <https://www.marinespecies.org/>

## What is `worrms`?

`worrms` is the official R client for the WoRMS REST API, developed by
[rOpenSci](https://ropensci.org/). It wraps every public API endpoint into
tidy, tibble-returning R functions.

**Key design principles:**

- Every function name begins with `wm_` (WoRMS marker)
- Functions ending in `_` accept **vectorised** inputs (multiple IDs / names in one call)
- All results are returned as `tibble's` or lists of `tibble's`

## Session goals

1. Install and load `worrms`
2. Walk through **every** function family with Antarctic examples
3. Demonstrate a full biodiversity use-case combining `worrms` with the **GBIF** occurrence data
   (via `rgbif`)

---

# Setup

## Installation

```{r install, eval=FALSE}
# From CRAN
install.packages("worrms")

# Or the development version from GitHub (requires remotes)
# remotes::install_github("ropensci/worrms")

# Companion packages used in this tutorial
install.packages(c("tidyverse", "rgbif", "sf", "rnaturalearth",
                   "rnaturalearthdata", "DT", "knitr", "kableExtra"))
```

## Load packages

```{r load-packages}
library(worrms)      # WoRMS API client
library(tidyverse)   # Data wrangling + ggplot2
library(rgbif)       # GBIF occurrence data
library(sf)          # Spatial operations
library(rnaturalearth)   # World basemap
library(rnaturalearthdata)
library(knitr)       # Tables in Rmd
library(kableExtra)  # Styled tables
```

> **API note:** `worrms` calls the WoRMS REST API over HTTPS — no authentication key required.
> Be mindful of rate limits: avoid looping thousands of single calls; use the vectorised
> `_` variants instead.

---

# Core concepts: the AphiaID

Every taxon in WoRMS has a unique integer identifier called the **AphiaID**. This is the
primary key used across all functions. The first thing you usually do is convert a
scientific name to an AphiaID.

> **Tip:** Always work with *accepted* AphiaIDs. `worrms` will tell you if a name is a
> synonym and what the valid name / ID is.

---

# Function family 1 — Name ↔ ID lookups

## `wm_name2id()` — Scientific name → AphiaID

The simplest entry point: give a name, get the ID.

```{r name2id}
# Single species: Antarctic krill
krill_id <- wm_name2id("Euphausia superba")
krill_id
```

```{r name2id-multi}
# Vectorised version: multiple names at once (note the trailing _)
antarctic_names <- c(
  "Euphausia superba",        # Antarctic krill
  "Dissostichus eleginoides", # Patagonian toothfish
  "Aptenodytes forsteri",     # Emperor penguin (marine but also terrestrial)
  "Notothenia coriiceps",     # Black rockcod
  "Balaenoptera musculus",    # Blue whale
  "Limacina helicina"         # Pteropod sea butterfly
)

aphia_ids <- wm_name2id_(antarctic_names)
aphia_ids   # Returns a named list
```

## `wm_id2name()` — AphiaID → Scientific name

Reverse lookup: useful when you have a list of IDs from another database.

```{r id2name}
wm_id2name(236217)   # single ID — Euphausia superba

# Vectorised version: look up IDs dynamically so they never go stale
id_vec <- unlist(wm_name2id_(c("Euphausia superba",
                                "Dissostichus eleginoides",
                                "Notothenia coriiceps")))
wm_id2name_(id_vec)
```

---

# Function family 2 — Full taxon records

## `wm_record()` — Complete AphiaRecord for a single ID

This is the workhorse function. Returns ~25 fields per taxon.

```{r wm-record}
krill_record <- wm_record(236217)   # Euphausia superba AphiaID

# Show all columns
glimpse(krill_record)
```

Key fields to know:

| Field | Meaning |
|---|---|
| `AphiaID` | Unique WoRMS identifier |
| `scientificname` | Accepted or queried name |
| `status` | `"accepted"`, `"unaccepted"` (synonym), etc. |
| `valid_AphiaID` | AphiaID of the accepted name (differs from AphiaID if synonym) |
| `rank` | Taxonomic rank (Species, Genus, …) |
| `kingdom` / `phylum` / `class` … | Higher taxonomy |
| `isMarine` / `isBrackish` / `isFreshwater` / `isTerrestrial` | Habitat flags |
| `isExtinct` | Extinction status |
| `citation` | How to cite this record |
| `lsid` | Life Science Identifier (globally unique URI) |

## `wm_records_name()` — Search by name (fuzzy optional)

```{r records-name}
# Exact match
wm_records_name("Euphausia superba") |> select(AphiaID, scientificname, status, rank)

# Fuzzy matching — great for typos or uncertain spelling
wm_records_name("Euphausia su", fuzzy = TRUE) |>
  select(AphiaID, scientificname, status, rank, match_type)
```

## `wm_records_names()` — Multiple names at once

```{r records-names}
# Vectorised version for a species list
multi_records <- wm_records_names(antarctic_names)

# Each element of the list is a tibble for one input name
names(multi_records) <- antarctic_names

# Bind them and keep key columns
all_records <- bind_rows(multi_records, .id = "query") |>
  select(query, AphiaID, scientificname, status, rank, isMarine, isExtinct)

kbl(all_records, caption = "WoRMS records for selected Antarctic taxa") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## `wm_records_taxamatch()` — TAXAMATCH fuzzy matching algorithm

TAXAMATCH is WoRMS' own name-matching algorithm — more sophisticated than simple fuzzy
matching. Ideal for cleaning up species lists from field data.

```{r taxamatch}
# Try some slightly messy names
messy_names <- c("Euphausia superba",     # correct
                 "Euphausia supurba",      # typo
                 "Notothenia coriceps")    # missing letter

taxamatch_results <- wm_records_taxamatch(messy_names)

bind_rows(taxamatch_results, .id = "query_index") |>
  select(query_index, scientificname, status, match_type)
```

## `wm_records_common()` — Search by vernacular (common) name

```{r records-common}
# Find species whose common name includes "krill"
krill_results <- wm_records_common("krill", fuzzy = FALSE)
krill_results |> select(AphiaID, scientificname, rank) |> head(10)
```

## `wm_records_rank()` — Browse by taxonomic rank within a group

```{r records-rank}
# Look up the AphiaID for Euphausiidae dynamically (avoids hardcoded IDs going stale)
euphausiidae_id <- wm_name2id("Euphausiidae")
cat("Euphausiidae AphiaID:", euphausiidae_id, "\n")

# All accepted species in Euphausiidae — rank 220 = Species in WoRMS
euphausiid_spp <- wm_records_rank(220, id = euphausiidae_id) |> distinct(scientificname, AphiaID, .keep_all = TRUE)
nrow(euphausiid_spp)
euphausiid_spp |> select(AphiaID, scientificname, status) |> head(10)
```

## `wm_records_date()` — Records modified since a given date

Useful for tracking updates to the database over time.

```{r records-date}
# Records added or modified since the start of 2024
recent <- wm_records_date("2024-01-01T00:00:00+00:00")
nrow(recent)
recent |> select(AphiaID, scientificname, rank, modified) |> head(8)
```

---

# Function family 3 — Taxonomy tree navigation

## `wm_classification()` — Full classification path

```{r classification}
# Full taxonomy from kingdom to species for Antarctic krill
krill_class <- wm_classification(236217)
krill_class

# Tidy view
krill_class |>
  select(rank, scientificname) |>
  kbl(caption = "Taxonomic classification of *Euphausia superba*") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## `wm_children()` — Direct taxonomic children

```{r children}
# Children of order Euphausiacea (AphiaID = 1128)
# We look this up dynamically first to future-proof the code
euphausiacea_id <- wm_name2id("Euphausiacea")
cat("Euphausiacea AphiaID:", euphausiacea_id, "\n")

# wm_children() returns HTTP 204 (No Content) — and therefore throws an error —
# when a taxon has no registered children. Always guard with tryCatch().
euphausiacea_children <- wm_children(euphausiacea_id)
euphausiacea_children |> select(AphiaID, scientificname, rank) |> head(15)
```

## `wm_ranks_id()` and `wm_ranks_name()` — Look up taxonomic rank names
**Important:**The same rank (e.g., Species) exists independently in multiple kingdoms

```{r ranks}
# What is rank 220?
wm_ranks_id(220)

# What rank ID does "Species" have?
wm_ranks_name("Species")
```

---

# Function family 4 — Synonyms & nomenclature

## `wm_synonyms()` — All synonyms for a taxon

```{r synonyms}
# Synonyms of Euphausia superba
syn <- wm_synonyms(236217)
if (nrow(syn) > 0) {
  syn |> select(AphiaID, scientificname, status, unacceptreason)
} else {
  cat("No synonyms found for this taxon.\n")
}
```

```{r synonyms-toothfish}
# Patagonian toothfish 
toothfish_id <- wm_name2id("Dissostichus eleginoides")
toothfish_syn <- wm_synonyms(toothfish_id)
toothfish_syn |> select(AphiaID, scientificname, status)
```

---

# Function family 5 — Vernacular names

## `wm_common_id()` — Common names for an AphiaID

```{r common-names}
common <- wm_common_id(236217)   # Euphausia superba
common

```

---

# Function family 6 — Geographic distribution

## `wm_distribution()` — Distribution records by AphiaID

```{r distribution}
krill_dist <- wm_distribution(236217)
glimpse(krill_dist)

# How many records per locality type?
krill_dist |> count(locality, sort = TRUE) |> head(15)
```

---

# Function family 7 — External identifiers

## `wm_external()` — Get an external ID for a taxon

WoRMS stores cross-references to many other databases. Common type codes:

| Code | Database |
|---|---|
| `tsn` | ITIS Taxonomic Serial Number |
| `bold` | Barcode of Life Database |
| `eol` | Encyclopedia of Life |
| `col` | Catalogue of Life |
| `iucn` | IUCN Red List ID |
| `fishbase` | Fishbase species key |

```{r external}
# Get the GBIF taxon key for Antarctic krill
wm_external(234700, type = "fishbase") # Patagonian toothfish

```

## `wm_record_by_external()` — Go the other way: external ID → WoRMS record

```{r record-by-external}
# Find WoRMS record using an ITIS TSN (Integrated Taxonomic Information System)
# TSN 96559 = Euphausia superba in ITIS
wm_record_by_external(96559, type = "tsn")
```

---

# Function family 8 — Sources & citations

## `wm_sources()` — Literature sources for a taxon

```{r sources}
sources <- wm_sources(236217)
sources |> select(source_id, reference, page) |> head(5)
```

---

# Function family 9 — Trait attributes

Attributes are the most powerful and underused part of WoRMS. They encode functional
traits like body size, diet, habitat, IUCN status, etc.

## `wm_attr_def()` — Attribute type definitions

```{r attr-def}
# Type 1 = IUCN Red List Category
wm_attr_def(id = 1)

# Type 15 = Body size
wm_attr_def(id = 15)
```

## `wm_attr_category()` — Attribute category values

```{r attr-category}
# What values exist for functional group (CategoryID = 7)?
func_groups <- wm_attr_category(id = 7)
func_groups |> select(measurementValueID, measurementValue)
```

## `wm_attr_data()` — Trait data for a specific taxon

```{r attr-data}
# Attributes for Antarctic krill
krill_attrs <- wm_attr_data(236217)
krill_attrs |> glimpse()
```

```{r attr-data-multi}
# Vectorised: get traits for several taxa at once
ids_to_check <- unlist(wm_name2id_(c("Euphausia superba",
                                      "Dissostichus eleginoides",
                                      "Notothenia coriiceps")))
attr_list <- wm_attr_data_(ids_to_check)

# Bind and summarise
attr_all <- bind_rows(attr_list)
attr_all |>
  select(AphiaID, measurementType, measurementValue) |>
  head(20)
```

## `wm_attr_aphia()` — All taxa that have a given attribute

```{r attr-aphia}
# Get AphiaIDs of taxa that have any attribute from category 1 (IUCN Red List)
# (returns up to 50 per page — the API is paginated)
taxa_with_fg <- wm_attr_aphia(id = 1)
head(taxa_with_fg, 10)
```

---

# Full use-case — Antarctic fish biodiversity with WoRMS + GBIF

Here we bring everything together. We will:

1. Identify the taxonomic subtree of Antarctic *Notothenioidei* (the dominant Antarctic fish group) via WoRMS
2. Resolve accepted names and AphiaIDs
3. Retrieve occurrence records from GBIF
4. Enrich with WoRMS trait data
5. Visualise species richness and trait composition

## Step 1 — Find the Notothenioidei subtree in WoRMS

```{r usecase-subtree}
# AphiaID for suborder Notothenioidei
notothenioidei_id <- wm_name2id("Notothenioidei")
notothenioidei_id

# Classification to confirm position
wm_classification(notothenioidei_id) |> select(rank, scientificname)
```

```{r usecase-families}
# Direct children (families) — look up ID dynamically
notothenioidei_id <- wm_name2id("Notothenioidei")

notothenioidei_families <- tryCatch(
  wm_children(notothenioidei_id),
  error = function(e) {
    message("wm_children() returned no content or errored: ", conditionMessage(e))
    tibble()
  }
)
notothenioidei_families |> select(AphiaID, scientificname, rank)
```

```{r usecase-species}
# For each family, get all species (rank 220)
# We use wm_records_rank_ (vectorised) where possible
family_ids <- notothenioidei_families |>
  filter(rank == "Family") |>
  pull(AphiaID)

# Retrieve species per family — wrap in purrr for safe iteration
species_list <- map(family_ids, function(fid) {
  tryCatch(
    wm_records_rank(220, id = fid),
    error = function(e) NULL
  )
})

# Name the list by family
names(species_list) <- notothenioidei_families |>
  filter(rank == "Family") |>
  pull(scientificname)

# Bind into one data frame
notothenioidei_spp <- bind_rows(species_list, .id = "family") |>
  filter(status == "accepted")

cat("Total accepted Notothenioidei species in WoRMS:", nrow(notothenioidei_spp), "\n")

notothenioidei_spp |>
  count(family, name = "n_species") |>
  arrange(desc(n_species)) |>
  kbl(caption = "Accepted species per Notothenioidei family") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## Step 2 — Retrieve GBIF occurrences south of 60°S

```{r usecase-gbif, eval=TRUE}
# Get accepted scientific names to pass to GBIF
target_species <- notothenioidei_spp |>
 # slice_head(n = 20) |>   # limit to 20 spp for demo (remove slice_head for full run)
  pull(scientificname)

# Download occurrences from GBIF via occ_search()
gbif_results <- map(target_species, function(sp) {
  tryCatch(
    occ_search(
      scientificName = sp,
      decimalLatitude = "-90,-60",   # south of 60°S
      hasCoordinate  = TRUE,
      limit          = 200, # limit to 200 records per species for demo
      fields         = c("name", "decimalLatitude", "decimalLongitude",
                         "year", "datasetName", "species")
    )$data,
    error = function(e) NULL
  )
})

names(gbif_results) <- target_species
gbif_df <- bind_rows(gbif_results) |>
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude))

cat("Total GBIF occurrence records retrieved:", nrow(gbif_df), "\n")
```

## Step 3 — Join WoRMS metadata onto GBIF records

```{r usecase-join}
# Get WoRMS AphiaIDs for the same species names
worms_lookup <- notothenioidei_spp |>
  filter(scientificname %in% target_species) |>
  select(scientificname, AphiaID, family, isMarine, isExtinct, citation)

# Join
gbif_enriched <- gbif_df |>
  rename(scientificname = species) |>
  left_join(worms_lookup, by = "scientificname")

glimpse(gbif_enriched)
```

## Step 4 — Retrieve WoRMS trait data and join

```{r usecase-traits}
# Get AphiaIDs
focal_ids <- worms_lookup$AphiaID

# Fetch trait data (may take a moment)
trait_list <- map(focal_ids, function(aid) {
  tryCatch(wm_attr_data(aid), error = function(e) NULL)
})
names(trait_list) <- focal_ids

trait_df <- bind_rows(trait_list, .id = "AphiaID") |>
  mutate(AphiaID = as.integer(AphiaID)) |>
  filter(measurementType %in% c("Functional group", "IUCN Red List Category",
                                 "Body size", "Diet"))

# Pivot wider for easy use
trait_wide <- trait_df |>
  select(AphiaID, measurementType, measurementValue) |>
  distinct() |>
  pivot_wider(names_from = measurementType, values_from = measurementValue,
              values_fn = \(x) paste(unique(x), collapse = "; "))

# Final enriched species table
species_enriched <- worms_lookup |>
  left_join(trait_wide, by = "AphiaID")

kbl(species_enriched |> select(scientificname,`Body size`),
    caption = "Notothenioidei species with WoRMS Body Size") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Step 5 — Visualise

### Occurrence map — south polar projection

```{r usecase-map, fig.cap="GBIF occurrences of Notothenioidei south of 60°S"}
world <- ne_countries(scale = "medium", returnclass = "sf")

gbif_sf <- st_as_sf(gbif_enriched,
                    coords = c("decimalLongitude", "decimalLatitude"),
                    crs = 4326)

ggplot() +
  geom_sf(data = st_transform(world, 3031),
          fill = "grey80", colour = "white", linewidth = 0.15) +
  geom_sf(data = st_transform(gbif_sf, 3031),
          aes(colour = family),
          alpha = 0.7, size = 1.8) +
  coord_sf(crs = 3031,
           xlim = c(-4000000, 4000000),
           ylim = c(-4000000, 4000000)) +
  scale_colour_brewer(palette = "Set3", name = "Family") +
  labs(title = "Notothenioidei occurrence records (GBIF, south of 60S)",
       subtitle = paste0(nrow(gbif_enriched), " records | ",
                         n_distinct(gbif_enriched$scientificname), " species"),
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold"))
```

### Species richness per family

```{r usecase-richness-plot, fig.cap="Species richness per Notothenioidei family"}
notothenioidei_spp |>
  count(family, name = "n_species") |>
  mutate(family = fct_reorder(family, n_species)) |>
  ggplot(aes(x = n_species, y = family, fill = n_species)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_c(option = "mako", direction = -1) +
  labs(title    = "Species richness across Notothenioidei families (WoRMS)",
       subtitle = "Accepted species only",
       x = "Number of accepted species", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

### Occurrence records per decade

```{r usecase-temporal, fig.cap="GBIF records per decade for focal Notothenioidei"}
gbif_enriched |>
  filter(!is.na(year)) |>
  mutate(decade = (year %/% 10) * 10) |>
  count(decade, family) |>
  ggplot(aes(x = decade, y = n, fill = family)) +
  geom_col() +
  scale_fill_brewer(palette = "Set3", name = "Family") +
  labs(title    = "GBIF sampling effort through time — Notothenioidei",
       x = "Decade", y = "Number of occurrence records") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold"))
```

---

# Quick-reference cheat sheet

```{r cheatsheet, echo=FALSE}
cheatsheet <- tibble::tribble(
  ~`Function`,               ~`Purpose`,                                        ~`Vectorised version`,
  "wm_name2id()",            "Scientific name → AphiaID",                       "wm_name2id_()",
  "wm_id2name()",            "AphiaID → scientific name",                       "wm_id2name_()",
  "wm_record()",             "Full AphiaRecord for one ID",                     "wm_record_()",
  "wm_records_name()",       "Search by name (fuzzy optional)",                 "—",
  "wm_records_names()",      "Multiple names → list of records",                "—",
  "wm_records_taxamatch()",  "TAXAMATCH fuzzy matching",                        "—",
  "wm_records_common()",     "Search by vernacular name",                       "wm_records_common_()",
  "wm_records_rank()",       "All taxa of a given rank under a parent",         "—",
  "wm_records_date()",       "Records modified since a date",                   "—",
  "wm_classification()",     "Full taxonomic classification",                   "wm_classification_()",
  "wm_children()",           "Direct taxonomic children",                       "wm_children_()",
  "wm_synonyms()",           "All synonyms / unaccepted names",                 "wm_synonyms_()",
  "wm_common_id()",          "Vernacular names for an AphiaID",                 "wm_common_id_()",
  "wm_distribution()",       "Geographic distribution records",                 "wm_distribution_()",
  "wm_external()",           "External database ID (GBIF, IUCN, …)",           "wm_external_()",
  "wm_record_by_external()", "External ID → WoRMS record",                     "wm_record_by_external_()",
  "wm_sources()",            "Literature sources / references",                 "wm_sources_()",
  "wm_attr_def()",           "Attribute type definitions",                      "wm_attr_def_()",
  "wm_attr_category()",      "Attribute category values",                       "wm_attr_category_()",
  "wm_attr_data()",          "Trait attribute data for a taxon",                "wm_attr_data_()",
  "wm_attr_aphia()",         "AphiaIDs with a given attribute",                 "wm_attr_aphia_()",
  "wm_ranks_id()",           "Rank name from rank ID",                          "—",
  "wm_ranks_name()",         "Rank ID from rank name",                          "—"
)

kbl(cheatsheet, caption = "worrms function cheat sheet") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE,
                font_size = 13) |>
  column_spec(1, monospace = TRUE, bold = TRUE, color = "#2C7BB6")
```

---

# Tips & best practices

**1. Always use vectorised `_` variants for multiple taxa**

```r
# Slow — one API call per species
results <- map(my_ids, wm_record)

# Fast — one API call for up to 50 IDs
results <- wm_record_(my_ids)
```

**2. Handle `NULL` / empty results gracefully**

The WoRMS API returns HTTP 204 (No Content) when nothing matches. `worrms` converts these
to `NULL` or empty tibbles. Always use `tryCatch()` or `purrr::possibly()` in loops.

```r
safe_record <- possibly(wm_record, otherwise = NULL)
results <- map(my_ids, safe_record)
```

**3. Check `status` before using records**

Some names are unaccepted synonyms. Always filter or redirect to `valid_AphiaID`.

```r
records |> filter(status == "accepted")
# or
records |> mutate(use_id = if_else(status == "accepted", AphiaID, valid_AphiaID))
```

**4. Cite WoRMS in your work**

Use `wm_sources()` to get the exact reference for each taxon, or cite the database itself:

> WoRMS Editorial Board (2024). World Register of Marine Species.
> Available from <https://www.marinespecies.org> at VLIZ.
> Accessed `r Sys.Date()`. DOI: 10.14284/170

---

# Session information

```{r session-info}
sessionInfo()
```
